A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-24 code/figure2.Rmd

```{r figure-2-plot, fig.width = 9}
umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv") ## load data
load("../data/kelly.colours.rda")
cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
names(cluster.colours) <- c(0:10) ## assigning cluster to the colors

@@

opar <- par(mfrow = c(1,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
par(mar = c(2,2,0,0), bg = NA) ## setting plot margins for the following plots

for(n in 1) {
  with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
                                  theta = theta[n], phi = phi[n], scale = T,
                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                  colvar = NULL))
  title(xlab="UMAP 2", line=-5.2)
  for(p in 2:length(phi)) {
    with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, 
                                theta = theta[p], phi = phi[p], scale = T,
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
  }
}

>>>>

```{r figure-2-plot, fig.width = 9, echo = FALSE}
umap_cancertype <- data.table::fread("../data/umap_3d_coors.tsv") ## load data
load("../data/kelly.colours.rda")
cluster.colours <- c("grey", kelly.colours[c(3:12)]) ## Color scheme for cluster
names(cluster.colours) <- c(0:10) ## assigning cluster to the colors

@@

opar <- par(mfrow = c(1,3), mai = c(0.1, 0.1, 0.1, 0.1)) ## plot layout 2x3
par(mar = c(0,2,0,0), bg = NA) ## setting plot margins for the following plots

for(n in 1) {
  with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
                                  bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, # clust_knn is knn result
                                  theta = theta[n], phi = phi[n], scale = T,
                                  xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                  colvar = NULL))
  title(xlab="UMAP 2", line=-5.2)
  for(p in 2:length(phi)) {
    with(umap_cancertype, scatter3D(plot_y, -plot_z, plot_x, 
                                bg = cluster.colours[as.character(clust_knn)], pch = 21, cex = 0.8, lwd = 0.2, 
                                theta = theta[p], phi = phi[p], scale = T,
                                xlab = "", ylab = "UMAP 3", zlab = "UMAP 1",
                                colvar = NULL))
        if(p == 2) {
      mtext("Cluster", side=3, line=-3.5, cex=1.2, adj=0.5)
      legend("top", inset = c(0.10), legend=names(cluster.colours[2:6]), col=cluster.colours[2:6], pch = 20, bty = "n", horiz = T, cex = 1.2)
      legend("top", inset = c(0.15), legend=names(cluster.colours[7:11]), col=cluster.colours[7:11], pch = 20, bty = "n", horiz = T, cex = 1.2)
    }
  }
}

### 2020-08-24 text/manuscript.Rmd

```{r, child='../code/figure2.Rmd'}
```

@@

**Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
a) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the
UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of
tumor stage onto the UMAP, regardless of cancer type. e) Projection of
metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
metastatic tumors.

@@

In order to visualize pathway disruption enrichment across all cancers
(pan-cancer), we created waterfall plots, replacing genes with pathways
(**figure 5A**). A complete list of pathway disruptions with percent mutated
samples and top genes is provided in **supplementary table ST1**. As expected,
these analyses reveal the broad importance of many well known pathways that are
disrupted in cancer, including "PIP3 activates Akt signaling" (77% of samples),
"MAP1K/MAP3K signaling" (70% of samples), "Mitotic G2-G2/M phases" (67% of
samples), "Cellular senescence" (64% of samples), "G2/M Checkpoints" (62% of
samples), _etc._

@@

**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
a) Waterfall plot; columns represent tumors. Each row represents a pathway. The
pathways are ranked such that the pathway with the most disruptions across all
7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
top pathway disruptions successively. The top 100 most mutated pathways are
shown. Red signifies pathway disrupted pathways, grey signifies not disrupted.
b) Heatmap shows the relative enrichment of each pathway (columns) within each
numbered cluster (rows). Effect size is displayed as colors (orange, enrichment;
purple, depletion). c) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size >= 0.30 in each
cluster) within each numbered cluster (rows).

@@

Wish wish to thank David Van Valen, Simon Knott and Kate Lawrenson for critical
reading of this manuscript, and Ivetth Corona for early discussions and
feedback.

>>>>

#### Figure 2
```{r, child='../code/figure2.Rmd'}
```

**Figure 2: Dimensionality reduction of tumor samples using pathway aberrations data 
identifies ten well-defined clusters.** Different perspectives of the same data in 
three-dimensional space are shown using UMAP. Each dot corresponds to a tumor sample. 
Colors indicate the tumor's cluster identity.

@@

**Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
a) Sankey diagram (flow is left to right) of the tumor type make up in a cluster. Cluster 
membership is on the left. Cancer type is on the right. b) Projection of breast cancer
subtypes onto the UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) 
Projection of tumor stage onto the UMAP, regardless of cancer type. e) Projection of
metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
metastatic tumors.

@@

In order to identify pathway disruption enrichment across all cancers
(pan-cancer), we created a list of pathway disruptions with percent mutated
samples and top genes (**supplementary table ST1**). As expected,
these analyses reveal the broad importance of many well known pathways that are
disrupted in cancer, including "PIP3 activates Akt signaling" (77% of samples),
"MAP1K/MAP3K signaling" (70% of samples), "Mitotic G2-G2/M phases" (67% of
samples), "Cellular senescence" (64% of samples), "G2/M Checkpoints" (62% of
samples), _etc._

@@

**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
a) Heatmap shows the relative enrichment of each pathway (columns) within each
numbered cluster (rows). Effect size is displayed as colors (orange, enrichment;
purple, depletion). b) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size >= 0.30 in each
cluster) within each numbered cluster (rows).

@@

We wish to thank David Van Valen, Simon Knott and Kate Lawrenson for critical
reading of this manuscript, and Ivetth Corona for early discussions and
feedback.
